<!-- LTeX: language=en-US -->
# Extending and Evaluating the Linear Model {#sec-extendLinRegCh}

We continue with the `wage1` data and the models fitted in
@sec-simpleLinRegCh and @sec-multipleLinRegCh. Since each chapter's R code
runs in its own fresh session, we reload the data and packages:

```{r}
library("wooldridge")
library("kableExtra")
suppressPackageStartupMessages(library("sandwich"))
suppressPackageStartupMessages(library("lmtest"))
data("wage1")

## Refit the models from @sec-simpleLinRegCh and @sec-multipleLinRegCh
## that we reuse below (model-fit comparisons, etc.)
lm_simple     <- lm(wage ~ educ,                 data = wage1)
lm_educ_exper <- lm(wage ~ educ + exper,         data = wage1)
lm_multiple   <- lm(wage ~ educ + exper + tenure, data = wage1)
```

## Extensions of the Linear Model {#sec-extensions}

The multiple linear regression model @eq-multipleLR is **linear in the
parameters** $\beta_1, \dots, \beta_K$, not necessarily in the original
variables. We can use linear and non-linear transformations such as 

* Transformed outcome: $\log(Y_i)$ 
* New predictor, for instance, 
    * by squaring: $X_{i4} = X_{i3}^2$
    * by interactions: $X_{i6} = \text{educ}_i \times \text{female}_i$ (@sec-interactions)

to generate new outcome and predictor variables. 

Note: The model remains linear in the parameters $\beta_1, \dots, \beta_K,$ so all results from @sec-multipleLinReg continue to apply.

### Log Transformations: The Mincer Equation {#sec-logtrans}

#### **Motivation: The Distribution of Wages** {-}

Hourly wages are right-skewed: a few high earners pull the mean well above
the median. Taking logarithms compresses the right tail and produces a
more symmetric distribution.

```{r}
#| label: fig-wage-hist
#| echo: false
#| fig-cap: >
#|   Distribution of hourly wages (left) and log hourly wages (right)
#|   for 526 workers. Taking logarithms reduces the skewness from
#|   2.0 to 0.4 and brings the distribution closer to symmetry.
par(mfrow = c(1, 2))
hist(wage1$wage,
     breaks = 25, col = "steelblue", border = "white",
     main = "Wage (USD/hour)",
     xlab = "Wage", las = 1)
hist(wage1$lwage,
     breaks = 25, col = "steelblue", border = "white",
     main = "Log Wage",
     xlab = "log(Wage)", las = 1)
par(mfrow = c(1, 1))
```

::: {.callout-note collapse="true"}
## R Code: Histograms of Wages and Log Wages
```{r}
#| ref.label: fig-wage-hist
#| eval: false
```
:::

#### **Interpretation of Log-Level Models** {-}

In a **log-level model**
$$
\log(Y_i) = \beta_1 + \beta_2 X_i + \varepsilon_i,
$$
a one-unit increase in $X_i$ changes the expected *log* of $Y_i$ by
$\beta_2$. Since differences in logs are approximately relative changes,
$$
\log(y_1) - \log(y_0) \approx \frac{y_1 - y_0}{y_0} \quad \text{for small changes,}
$$

* a one-unit increase in $X_i$ is associated with an **approximate** $100 \cdot \beta_2\%$ change in $Y_i$;
<!-- 
* the **exact** change is $100 \cdot (\exp(\beta_2) - 1)\%$, which matters for larger $\beta_2$. 
-->

#### **The Mincer Earnings Equation** {-}

The standard specification in the economics of education is the
**Mincer equation** [@Mincer1974]:

$$
\log(\text{wage}_i)
= \beta_1 + \beta_2\,\text{educ}_i + \beta_3\,\text{exper}_i
+ \beta_4\,\text{exper}_i^2 + \varepsilon_i.
$${#eq-mincer}

Compared to the levels models in @sec-multipleLinReg, this specification:

1. Accommodates the right-skewed wage distribution.
2. Gives $\widehat\beta_2$ a natural interpretation as the **proportional
   return to one year of schooling**.
3. Includes a quadratic in experience to capture diminishing returns to
   experience (see below).

```{r}
lm_mincer <- lm(lwage ~ educ + exper + expersq, data = wage1)
coeftest(lm_mincer, vcov = vcovHC(lm_mincer, type = "HC3"))
```

**Interpretation of $\widehat\beta_2$:** Each additional year of education is
associated with approximately a
`r sprintf("%.1f", 100 * coef(lm_mincer)["educ"])`% increase in hourly wages
(exact: `r round(100 * (exp(coef(lm_mincer)["educ"]) - 1), 1)`%) — a much more
interpretable and economically meaningful number than the
`r round(100 * coef(lm_simple)["educ"])` cents per hour from the simple levels model.

#### **Polynomial Terms: Diminishing Returns to Experience** {-}

Including $\text{exper}_i^2$ turns @eq-mincer into a **polynomial
(quadratic) regression** in experience:

\begin{align*}
%Y_i
%&= \beta_1 + \beta_2\,X_{i2} + \beta_3\,X_{i3}
%+ \beta_4\,\overbrace{X_{i4}}^{\text{new predictor}} + %\varepsilon_i\\[2ex]
\log(\text{wage}_i)
&= \beta_1 + \beta_2\,\text{educ}_i + \beta_3\,\text{exper}_i
+ \beta_4\,\underbrace{\text{exper}_i^2}_{\text{new predictor}} + \varepsilon_i.
\end{align*}

::: {.callout-note appearance="simple"}
The above model is quadratic in $\text{exper}_i$, but still **linear in the parameters** $\beta_1, \dots, \beta_4.$

The squared `exper`-predictor is simply treated as a further (here fourth) predictor 
$$
X_{i4} = \text{exper}_i^2
$$ 
of a multiple linear regression model. 

All OLS theory from @sec-multipleLinReg therefore still applies.
:::

The effect of experience is then no longer constant; the **marginal
effect** of one more year depends on how much experience a worker already
has:

$$
\frac{\partial}{\partial\, \text{exper}} E\bigl(\log(\text{wage}_i) \mid \text{educ}_i, \text{exper}_i\bigr)
= \beta_3 + 2\beta_4\, \text{exper}_i.
$${#eq-marginal-poly}

With $\beta_3 > 0$ and $\beta_4 < 0$, wages rise with experience at a
decreasing rate and peak where the marginal effect is zero, at
$$
\text{exper}^* = -\beta_3/(2\beta_4).
$$

```{r}
#| label: fig-mincer-exper
#| echo: false
#| fig-cap: >
#|   Fitted Mincer wage-experience profile at mean education, back-transformed
#|   to USD/hour (left), and the marginal effect of one additional year of
#|   experience in percent (right).
b          <- coef(lm_mincer)
x_seq      <- seq(0, 50, length.out = 300)
log_fit    <- b["(Intercept)"] + b["educ"] * mean(wage1$educ) +
              b["exper"] * x_seq + b["expersq"] * x_seq^2
me         <- b["exper"] + 2 * b["expersq"] * x_seq
peak_exper <- -b["exper"] / (2 * b["expersq"])

par(mfrow = c(1, 2))
plot(x_seq, exp(log_fit),
     type = "l", col = "darkred", lwd = 2, las = 1,
     xlab = "Experience (years)", ylab = "Predicted Wage (USD/hour)")
abline(v = peak_exper, lty = 2, col = "grey60")
plot(x_seq, 100 * me,
     type = "l", col = "darkred", lwd = 2, las = 1,
     xlab = "Experience (years)", ylab = "Marginal effect (% per year)")
abline(h = 0, lty = 2, col = "grey60")
par(mfrow = c(1, 1))
```

::: {.callout-note collapse="true"}
## R Code: Plotting the Mincer Experience Profile
```{r}
#| ref.label: fig-mincer-exper
#| eval: false
```
:::

**Interpretation:**

* First year of experience: about `r round(100 * b["exper"], 1)`% higher wages.
* A year after 20 years of experience: only about
  `r round(100 * (b["exper"] + 2 * b["expersq"] * 20), 1)`%.
* Wages peak at approximately `r round(peak_exper)` years of experience.

This is consistent with the Mincer literature: wages rise steeply early in
a career and level off later.


## Detecting Potential Problems {#sec-diagnostics}

Before interpreting or reporting a fitted model, we should check whether
the model assumptions are plausible for the data at hand. Three
residual-based diagnostic plots catch the most common violations.

We illustrate using the Mincer equation `lm_mincer` from @sec-logtrans.
The `plot()` method for `lm` objects produces these plots automatically:

```{r}
#| label: fig-diagnostics
#| echo: false
#| fig-cap: >
#|   Residual diagnostic plots for the Mincer equation
#|   log(wage) ~ educ + exper + exper². Panels: (1) residuals vs fitted,
#|   (2) normal Q-Q, (3) scale-location.
par(mfrow = c(1, 3))
plot(lm_mincer, which = 1, sub.caption = "log(wage) model")
plot(lm_mincer, which = 2, sub.caption = "log(wage) model")
plot(lm_mincer, which = 3, sub.caption = "log(wage) model")
par(mfrow = c(1, 1))
```

::: {.callout-note collapse="true"}
## R Code: Residual Diagnostic Plots
```{r}
#| ref.label: fig-diagnostics
#| eval: false
```
:::

#### **Plot 1: Residuals vs. Fitted Values** {-}

```{r}
#| label: fig-diag-resfit
#| echo: false
#| fig-cap: >
#|   Residuals vs. fitted values for the Mincer equation
#|   log(wage) ~ educ + exper + exper².
plot(lm_mincer, which = 1, sub.caption = "log(wage) model")
```

::: {.callout-note collapse="true"}
## R Code: Residuals vs. Fitted Values
```{r}
#| ref.label: fig-diag-resfit
#| eval: false
```
:::

**What to look for:**

* Residuals scatter randomly around zero, no pattern → OK.
* Residuals show a systematic pattern → functional form misspecified (e.g., a missing quadratic term).
* Spread of residuals changes with fitted values →
  **heteroskedasticity**.

**In our model:** No strong pattern. The log transformation and
experience quadratic have captured the main non-linearities. Some spread variation remains → **heteroskedasticity**. Residuals at small fitted values $\widehat{Y}_i$ sit strictly above the zero line → **the model may be misspecified for this part of the data.**

#### **Plot 2: Normal Q-Q** {-}

```{r}
#| label: fig-diag-qq
#| echo: false
#| fig-cap: >
#|   Normal Q-Q plot of the standardized residuals for the Mincer equation
#|   log(wage) ~ educ + exper + exper².
plot(lm_mincer, which = 2, sub.caption = "log(wage) model")
```

::: {.callout-note collapse="true"}
## R Code: Normal Q-Q Plot
```{r}
#| ref.label: fig-diag-qq
#| eval: false
```
:::

**What to look for:** Points should fall close to the 45° line. Clear deviations from the 45° line indicate non-normal errors.

* Normality is **not** needed for the OLS estimators to be unbiased.
* But the $t$- and $F$-distributions in @eq-t-dist and @eq-F-stat are
  exact only if $\varepsilon_i$ is normally distributed.
* For large $n$, the **Central Limit Theorem** ensures approximate
  validity even without normality of $\varepsilon_i.$ Therefore, departures matter most in small samples.
* Exception: the prediction interval in @sec-prediction relies on normality of $\varepsilon_\text{new}$ even in large samples.

**In our model:** The residuals follow the line closely over most of the range. Only the lower tail shows some deviations (e.g., observation 24). With $n = `r nrow(wage1)`$, the Central Limit Theorem provides adequate coverage, and the inference results remain reliable, when based on HC3-robust standard errors.

#### **Plot 3: Scale-Location** {-}

```{r}
#| label: fig-diag-scaleloc
#| echo: false
#| fig-cap: >
#|   Scale-location plot for the Mincer equation
#|   log(wage) ~ educ + exper + exper².
plot(lm_mincer, which = 3, sub.caption = "log(wage) model")
```

::: {.callout-note collapse="true"}
## R Code: Scale-Location Plot
```{r}
#| ref.label: fig-diag-scaleloc
#| eval: false
```
:::

**What to look for:**

* The square root of the absolute standardized residuals should be roughly flat across fitted values. A systematic pattern → **heteroskedasticity**.
* Standardized residuals larger than 2 (in absolute value) → possible
  outliers, which should be investigated (e.g., a typo in the data).

**In our model:** A modest upward trend, i.e., some heteroskedasticity
remains even after the log transformation. This reinforces the
recommendation from @sec-robustSE: **always use HC3 robust standard
errors**. Observation 24 — already noted in Plot 2 — has the largest
standardized residual (in absolute value) in the whole sample and is thus
a possible outlier worth a closer look.

::: {.callout-note appearance="simple"}
Diagnostic plots reveal problems; they do not fix them! 

The appropriate response depends on the violation: 

* heteroskedasticity → robust SEs
* misspecified model → transform predictors or add polynomial terms
* outliers → investigate individual observations
:::


## Prediction vs. Inference {#sec-prediction}

Linear regression serves two distinct purposes, each with its own target
and its own measure of uncertainty:

::: {.spaced-list}
* **Inference on parameters:** *By how much does education raise wages
  on average?*
   * Target: E.g. the population parameter $\beta_2$.
   * Relevant uncertainty: $\widehat{\operatorname{SE}}(\widehat\beta_2 \mid X)$.
* **Prediction:** *What wage would we expect for a specific new worker?*
   * Target: the wage $Y_\text{new}$ of a new worker with predictor
     values $x_{\text{new},2}, \dots, x_{\text{new},K}$.
   * Relevant uncertainty: combines the uncertainties of the estimators $\widehat\beta_1, \dots,\widehat\beta_K$ and the inherent randomness of
     $Y_\text{new}$ around its conditional mean $m(x_\text{new}),$ due to new error term $\varepsilon_\text{new}.$
:::

### Assessing Model Fit: $R^2$ and Adjusted $R^2$ {#sec-model-fit}

Before we quantify the uncertainty of individual predictions, a first question is how well an estimated model predicts all the *observed* outcome values $Y_1,\dots,Y_n$ in our sample.

Recall the **fitted values**
\begin{align*}
\widehat{Y}_i 
& = \widehat{m}(x_i)\\[2ex]
& = \widehat\beta_1 + \widehat\beta_2\, X_{i2} + \dots + \widehat\beta_K\, X_{iK}.
\end{align*}

and **residuals**
$$
\widehat{\varepsilon}_i = Y_i - \widehat{Y}_i,
$$
from @sec-OLS-simple.


We could ask: How well does education alone, i.e., our simple linear regression model in @eq-simpleLR  
$$
Y_i = \beta_1 + \beta_2 \underbrace{X_i}_{=\texttt{educ}_i} + \varepsilon_i, \quad i = 1, \dots, n,
$$
predict the observed wages $Y_1,\dots,Y_n$? 


We measure this by decomposing the total variation in $Y_1,\dots,Y_n$ into an explained and an unexplained (residual) part:

$$
\underbrace{\sum_{i=1}^{n}(Y_i - \bar Y)^2}_{\text{TSS}}
=
\underbrace{\sum_{i=1}^{n}(\widehat Y_i - \bar Y)^2}_{\text{ESS}}
+
\underbrace{\sum_{i=1}^{n} \widehat{\varepsilon}_i^2}_{\text{RSS}}
$${#eq-SS-decomp}

\begin{align*}
\Leftrightarrow\;
\underbrace{\frac{1}{n}\sum_{i=1}^{n}(Y_i - \bar Y)^2}_{\text{Variance of $Y_1,\dots,Y_n$}}
&= \underbrace{\frac{1}{n}\sum_{i=1}^{n}(\widehat Y_i - \bar Y)^2}_{\text{Variance of $\widehat{Y}_1,\dots,\widehat{Y}_n$}}
+ \underbrace{\frac{1}{n}\sum_{i=1}^{n} \widehat{\varepsilon}_i^2}_{\text{Variance of $\widehat{\varepsilon}_1,\dots,\widehat{\varepsilon}_n$}},
\end{align*}


where

::: {.spaced-list}
* **TSS** is the Total Sum of Squares,
* **ESS** is the Explained Sum of Squares, and
* **RSS** is the Residual Sum of Squares.
:::

The **coefficient of determination** $R^2$ is the fraction of total
variation explained by the model:

$$
R^2 = \frac{\text{ESS}}{\text{TSS}} = 1 - \frac{\text{RSS}}{\text{TSS}},
\qquad 0 \leq R^2 \leq 1.
$${#eq-Rsquared}

```{r}
RSS <- sum(residuals(lm_simple)^2)
TSS <- sum((wage1$wage - mean(wage1$wage))^2)
R2  <- 1 - RSS / TSS
c(R2_by_hand = R2,
  R2_by_R    = summary(lm_simple)$r.squared)
```

**Interpretation:** $R^2 = `r round(R2, 3)`$ means that the simple linear regression model with the predictor education explains about
`r round(100 * R2, 1)`% of the variation in the observed wages $Y_1,\dots,Y_n.$ That is, education
is informative about wages, but many other factors matter too.

#### **Comparing the Model Fits of Models with Different Numbers of Predictors** {-}

$R^2$ from @eq-Rsquared **never decreases** when a predictor is added,
even if that predictor is pure noise. Comparing $R^2$ across models with
different numbers of predictors is therefore misleading.

**Solution:** The **adjusted $R^2$** penalizes for large numbers of parameters $K$:

$$
\bar R^2 = 1 - \frac{\text{RSS}/(n-K)}{\text{TSS}/(n-1)}
         = 1 - (1 - R^2)\frac{n-1}{n-K}.
$${#eq-adjR2}

In fact, $\bar R^2$ can decrease when adding poor predictors.

```{r}
#| label: tbl-model-comparison
#| echo: false
#| tbl-cap: "Coefficient estimates and fit statistics across nested models ($n = 526$)."
## compare three different models (lm_simple: K=2, lm_educ_exper: K=3, lm_multiple: K=4;
## all three refitted at the top of this chapter)

## Row names of the coefficient block. Indexing coef(...) by these names
## automatically inserts NA for coefficients a model does not contain.
nms <- c("(Intercept)", "educ", "exper", "tenure")

coef_tab <- cbind(coef(lm_simple)[nms],
                   coef(lm_educ_exper)[nms],
                   coef(lm_multiple)[nms])
coef_tab <- round(coef_tab, 3)
rownames(coef_tab) <- nms

fit_tab <- rbind(
  R2    = c(summary(lm_simple)$r.squared,
            summary(lm_educ_exper)$r.squared,
            summary(lm_multiple)$r.squared),
  AdjR2 = c(summary(lm_simple)$adj.r.squared,
            summary(lm_educ_exper)$adj.r.squared,
            summary(lm_multiple)$adj.r.squared)
)
fit_tab <- round(fit_tab, 3)

tab <- rbind(coef_tab, fit_tab)
colnames(tab) <- c("(1)", "(2)", "(3)")

options(knitr.kable.NA = "—")
kbl(tab, align = "r") |>
  kable_styling(bootstrap_options = c("striped", "hover")) |>
  pack_rows("Coefficients", 1, 4) |>
  pack_rows("Model Fit Statistics", 5, 6)
```

::: {.callout-note collapse="true"}
## R Code: Table of Coefficient Estimates and Fit Statistics
```{r}
#| ref.label: tbl-model-comparison
#| eval: false
```
:::

* Adding experience raises $R^2$ from
`r sprintf("%.3f", summary(lm_simple)$r.squared)` to
`r sprintf("%.3f", summary(lm_educ_exper)$r.squared)`, and adding tenure raises it further to
`r sprintf("%.3f", summary(lm_multiple)$r.squared)`. 
* The adjusted $R^2$ rises as well.

::: {.callout-note appearance="simple"}
**A further note on $R^2$ comparisons.** 

The Mincer model with $\log(Y_i)$ outcomes  
`lm(lwage ~ educ + exper + expersq)`
from @sec-logtrans has
$R^2 = `r sprintf("%.2f", summary(lm_mincer)$r.squared)`,$ while the levels model
`lm(wage ~ educ + exper + expersq)` has
$R^2 = `r sprintf("%.2f", summary(lm(wage ~ educ + exper + expersq, data = wage1))$r.squared)`.$

‼️ These two $R^2$-statistics are **not directly comparable**, even though the predictors in both models are exactly the same. The reason: their outcome variables differ, $\log(Y_i)=\log(\text{wage}_i)$ vs. $Y_i=\text{wage}_i,$ so the TSS in the denominator of $R^2$ differs too. We would be comparing apples and oranges.
:::

<!-- 
### Confidence and Prediction Intervals {#sec-pred-intervals}

In both cases,

::: {.spaced-list}
* ***inference* on the conditional mean**
  \begin{align*}
  m(x_\text{new}) & = E(Y_\text{new} \mid x_\text{new})\\[2ex] &= \beta_1 + \beta_2\, x_{\text{new},2} + \dots + \beta_K\, x_{\text{new},K}
  \end{align*}
  and
* ***prediction* of a new outcome**
  $$
  Y_\text{new} = m(x_\text{new}) + \varepsilon_\text{new},
  $$
:::
the point estimate is the **fitted value**

\begin{align*}
\widehat{Y}_\text{new} 
& = \widehat{m}(x_\text{new})\\[2ex]
& = \widehat\beta_1 + \widehat\beta_2\, x_{\text{new},2} + \dots + \widehat\beta_K\, x_{\text{new},K}.
\end{align*}

However, to quantify the **uncertainties**, we need different interval estimates taking into account the different targets: 

::: {.spaced-list}
* **Confidence Intervals** quantify the uncertainty of the conditional mean estimator $\widehat{m}(x_\text{new}),$ which tries estimating the target $m(x_\text{new}).$
* **Prediction Intervals** quantify the uncertainties of the conditional mean estimatior $\widehat{m}(x_\text{new})$ plus the uncertainity due to the irreducable prediction error $\varepsilon_{\text{new}},$ which is part of the prediction target $Y_{\text{new}}.$
::: 

-->

### Confidence Intervals for the Conditional Mean Function {#sec-confMeanFun-intervals}

The conditional mean
\begin{align*}
  m(x_\text{new}) & = E(Y_\text{new} \mid x_\text{new})\\[2ex] &= \beta_1 + \beta_2\, x_{\text{new},2} + \dots + \beta_K\, x_{\text{new},K}
\end{align*}
i.e., the *average* wage of all workers with characteristics
$$
x_\text{new}=(x_{\text{new},2},\, \dots,\, x_{\text{new},K})
$$
is estimated by
$$
\widehat{m}(x_\text{new}) = \widehat\beta_1 + \widehat\beta_2\, x_{\text{new},2} + \dots + \widehat\beta_K\, x_{\text{new},K}.
$$

The estimator $\widehat{m}(x_\text{new})$ varies around the unknown true value $m(x_\text{new})$ from sample to sample because the estimated coefficients $\widehat{\beta}_1,\dots,\widehat{\beta}_K$ do.

For the **simple regression model**,
$$
Y_i=\underbrace{\beta_1+\beta_2X_i}_{=m(X_i)}+\varepsilon_i,
$$ 
the variance of $\widehat{m}(x_\text{new})$ is
$$
\operatorname{Var}(\widehat{m}(x_\text{new}) \mid X)
= \sigma^2 \left( \frac{1}{n} + \frac{(x_\text{new} - \bar X)^2}{\sum_{i=1}^{n}(X_i - \bar X)^2} \right).
$${#eq-CI-mean-var}

That is:
* This variance is smallest at $x_\text{new} = \bar X$ and grows as $x_\text{new}$ moves away from the center of the data. This is why the confidence band in @fig-wage-educ fans out at both ends.
* The variance shrinks as the sample size $n$ grows.

The variance formula in @eq-CI-mean-var cannot be computed directly, since we usually do not know $\sigma^2=\operatorname{Var}(\varepsilon_i).$ Under homoskedasticity, we can estimate $\sigma^2$ by $\widehat\sigma^2 = \frac{1}{n-K}\sum_{i=1}^n\widehat\varepsilon_i^2,$ which yields the feasible estimator
$$
\widehat{\operatorname{Var}}(\widehat{m}(x_\text{new}) \mid X)
= \widehat\sigma^2 \left( \frac{1}{n} + \frac{(x_\text{new} - \bar X)^2}{\sum_{i=1}^{n}(X_i - \bar X)^2} \right).
$$


For the **multiple regression model**,
$$
Y_i=\underbrace{\beta_1+\beta_2X_{i2} + \dots + \beta_K X_{iK}}_{m(X_i)} + \varepsilon_i,
$$
the formula for 
$$
\widehat{\operatorname{Var}}(\widehat{m}(x_\text{new}) \mid X)
$$
is longer but the logic is the same. 


The $(1-\alpha)\cdot 100\%$ **confidence interval for the conditional mean** is given by

$$
\widehat{m}(x_\text{new}) \;\pm\;
t_{n-K,\,1-\alpha/2}\cdot
\sqrt{\widehat{\operatorname{Var}}(\widehat{m}(x_\text{new}) \mid X)}.
$${#eq-CI-MeanFun-mean}

### Confidence Intervals for a Single Parameter {#sec-conf-intervals}

This is the same idea as @eq-CI-beta2 for the simple regression model, now generalized to $K$ parameters. A $(1-\alpha)\cdot 100\%$ **confidence interval for a single coefficient** $\beta_k$ is

$$
\widehat{\beta}_k \;\pm\;
t_{n-K,\,1-\alpha/2}\cdot
\widehat{\operatorname{SE}}(\widehat\beta_k \mid X).
$${#eq-CI-parameter}






### Prediction Interval for a New Observation {#sec-pred-intervals}

The second target is a *new individual outcome*
$Y_\text{new} = m(x_\text{new}) + \varepsilon_\text{new}$.
Its prediction error is

$$
Y_\text{new} - \widehat Y_\text{new}
= \underbrace{m(x_\text{new}) - \widehat Y_\text{new}}_{\text{estimation error}}
+ \underbrace{\varepsilon_\text{new}}_{\text{irreducible error}}.
$$

The new worker's error $\varepsilon_\text{new}$ is independent of the
sample used to estimate the coefficients, so the two variances add up:

$$
\operatorname{Var}(Y_\text{new} - \widehat Y_\text{new} \mid X)
= \sigma^2 + \operatorname{Var}(\widehat Y_\text{new} \mid X).
$${#eq-PI-var}

* The extra $\sigma^2$ in @eq-PI-var — absent from @eq-CI-mean-var —
  reflects the irreducible randomness of the individual outcome.
* A $95\%$ **prediction interval** is therefore always **wider** than the
  confidence interval for the mean.
* Unlike the confidence interval, the prediction interval additionally
  relies on the errors being (approximately) normally distributed — the
  Central Limit Theorem does not help here, since $\varepsilon_\text{new}$
  is a single draw, not an average:

$$
\widehat Y_\text{new} \;\pm\;
t_{n-K,\,0.975}\cdot
\sqrt{\widehat\sigma^2 + \widehat{\operatorname{SE}}(\widehat Y_\text{new} \mid X)^2}.
$${#eq-PI}

::: {.callout-tip collapse="true"}
## Going Deeper: The Intervals in Matrix Form

This box requires matrix algebra; see @sec-matrixAlgebra.

With $x_\text{new} = (1, x_{\text{new},2}, \dots, x_{\text{new},K})^\top$
and the design matrix $X$ from @sec-OLS-multiple, the fitted value is
$\widehat Y_\text{new} = x_\text{new}^\top\widehat\beta$ and

$$
\operatorname{Var}(\widehat Y_\text{new} \mid X)
= \sigma^2\, x_\text{new}^\top (X^\top X)^{-1} x_\text{new}.
$$

For the simple regression, this expression reduces to @eq-CI-mean-var.
:::


#### **Application: Benchmarking a Salary Offer** {-}

An HR department plans to hire a college graduate (16 years of education)
with 5 years of work experience and wants to know which hourly wage is
in line with the market. We use the Mincer model from @sec-logtrans and
transform the predictions back from log wages to dollars:

```{r}
new_worker <- data.frame(educ = 16, exper = 5, expersq = 25)

ci <- predict(lm_mincer, newdata = new_worker, interval = "confidence")
pi <- predict(lm_mincer, newdata = new_worker, interval = "prediction")

cat("Point estimate (USD/hour):             ", round(exp(ci[,"fit"]), 2), "\n")
cat("95% CI for mean wage (USD/hour):  [",
    round(exp(ci[,"lwr"]), 2), ",", round(exp(ci[,"upr"]), 2), "]\n")
cat("95% PI for individual wage (USD/hour): [",
    round(exp(pi[,"lwr"]), 2), ",", round(exp(pi[,"upr"]), 2), "]\n")
```

The two intervals answer two different HR questions:

* **"What do workers with this profile typically earn?"** The confidence
  interval [\$`r round(exp(ci[, "lwr"]), 2)`, \$`r round(exp(ci[, "upr"]), 2)`] is narrow: we can
  estimate the typical wage of *all* workers with this profile quite
  precisely.^[Strictly speaking, transforming back from log wages gives the
  *median* rather than the mean wage of workers with this profile.] This is
  the relevant benchmark when budgeting salaries for many similar hires.
* **"What might this particular candidate earn elsewhere?"** The prediction
  interval [\$`r round(exp(pi[, "lwr"]), 2)`, \$`r round(exp(pi[, "upr"]), 2)`] is about
  `r round(diff(exp(pi[, c("lwr", "upr")])) / diff(exp(ci[, c("lwr", "upr")])))` times as wide:
  predicting *one individual's* wage is far less certain, because wages
  vary enormously among workers with identical observable characteristics.
  This is the relevant range for negotiating an individual offer.

::: {.callout-note appearance="simple"}
As $n \to \infty$, the CI width $\to 0$ — we eventually learn $\beta$
exactly. The PI width does **not** shrink to zero: the irreducible
$\sigma^2$ term in @eq-PI-var remains even in infinite samples. No matter
how much data we have, an individual's wage contains randomness that no
model can eliminate.
:::

The figure below visualizes both bands along the education dimension,
holding `exper` at its sample mean:

```{r}
#| label: fig-CI-PI
#| echo: false
#| fig-cap: >
#|   95% confidence band for the conditional mean (dark red, narrow) and
#|   95% prediction band (salmon, wide) from the Mincer model, plotted
#|   against education at mean experience.
educ_seq  <- seq(0, 18, length.out = 200)
exper_bar <- mean(wage1$exper)
nd        <- data.frame(educ    = educ_seq,
                        exper   = exper_bar,
                        expersq = exper_bar^2)
ci_band <- predict(lm_mincer, newdata = nd, interval = "confidence")
pi_band <- predict(lm_mincer, newdata = nd, interval = "prediction")

plot(wage1$educ, wage1$lwage,
     col  = adjustcolor("steelblue", alpha.f = 0.3),
     pch  = 16, las = 1,
     xlab = "Education (years)", ylab = "Log Wage")
polygon(c(educ_seq, rev(educ_seq)),
        c(pi_band[,"lwr"], rev(pi_band[,"upr"])),
        col = adjustcolor("lightcoral", alpha.f = 0.25), border = NA)
polygon(c(educ_seq, rev(educ_seq)),
        c(ci_band[,"lwr"], rev(ci_band[,"upr"])),
        col = adjustcolor("darkred", alpha.f = 0.35), border = NA)
lines(educ_seq, ci_band[,"fit"], col = "darkred", lwd = 2)
legend("topleft",
       legend = c("Fitted mean", "95% CI (mean)", "95% PI (individual)"),
       col    = c("darkred",
                  adjustcolor("darkred",     alpha.f = 0.5),
                  adjustcolor("lightcoral",  alpha.f = 0.5)),
       lwd    = c(2, 8, 8), bty = "n")
```

::: {.callout-note collapse="true"}
## R Code: Plotting the Confidence and Prediction Bands
```{r}
#| ref.label: fig-CI-PI
#| eval: false
```
:::


## What Does $\widehat\beta_2$ Really Tell Us? {#sec-whatdoesbeta}

We now have a complete regression toolkit. The Mincer equation from
@sec-logtrans estimates that each additional year of education is associated
with approximately a **`r round(100 * coef(lm_mincer)["educ"])`% increase in hourly wages**, controlling for
experience. The inference tools from @sec-multipleLR-inference tell us this
estimate is statistically precise, and the model fits the data reasonably
well ($R^2 = `r sprintf("%.2f", summary(lm_mincer)$r.squared)`$).

So — does education make workers more productive and therefore better
paid? Can a student reading this script expect a `r round(100 * coef(lm_mincer)["educ"])`% wage premium for each
additional year they spend in university?

**Not necessarily.** The OLS estimate $\widehat\beta_\text{educ} = `r round(coef(lm_mincer)["educ"], 2)`$
measures a *statistical association*, not a causal effect. To see why,
consider two scenarios that are observationally indistinguishable with
our data:

* **Scenario A — education is causal:** More education raises
  productivity, which raises wages. The
  `r round(100 * coef(lm_mincer)["educ"])`% estimate reflects the true
  causal return.
* **Scenario B — ability bias:** Some workers are more able — they learn
  faster, work harder, think more creatively. High-ability workers both
  *choose* more education (less costly or more rewarding for them) *and*
  *earn higher wages* (for reasons unrelated to education). The regression
  then picks up the combined effect of education *and* the underlying
  ability differences between workers.

Under Scenario B, the omitted variable `ability` is correlated with both

* `educ` — more able workers acquire more education, and
* `wage` — more able workers earn more,

so $E(\varepsilon_i \mid \text{educ}_i) \neq 0$ — **Assumption 2
(exogeneity) is violated** — and, by the OVB formula @eq-OVB,

$$
E(\widehat\beta_\text{educ} \mid X) = \beta_\text{educ}^\text{causal}
+ \underbrace{\beta_\text{ability} \cdot \widehat\delta_\text{educ}}_{\text{ability bias} > 0}.
$$

* This argument suggests OLS *overstates* the causal return to education.
* Surprisingly, studies exploiting natural experiments — e.g., compulsory
  schooling laws that forced some students to stay in school longer —
  typically find causal returns **as large as or even larger than** the
  OLS estimates [@Card1999].
* Ability bias is not the only problem: errors in reported years of
  education bias OLS *downward*, and returns to education may differ
  across people.

**Lesson:** the direction and size of the bias are empirical questions
that OLS alone cannot answer.

**Prediction vs. causation.** Crucially, none of this undermines the
model's *predictive* usefulness. Even if $\widehat\beta_\text{educ}$ is
biased as a causal estimate, the Mincer equation still produces good
predictions of wages for workers similar to those in the sample. The
confidence and prediction intervals in @sec-prediction remain valid.
Bias and prediction accuracy are distinct properties of an estimator.

::: {.callout-important appearance="simple"}
## Business Takeaway: Does the Decision Need a Prediction or a Causal Effect?
Before acting on a regression, ask what kind of answer the decision
requires.

* *Budgeting salaries for next year's hires* needs a good **prediction** —
  OLS is well suited, even if $\widehat\beta_\text{educ}$ is not causal.
* *Deciding whether to fund employees' degree programs* needs the **causal
  effect** of education on wages (or productivity) — here, the OLS
  estimate may be misleading.
:::

::: {.callout-note appearance="simple"}
## What We Have and What We Are Missing

| | **Prediction** | **Causal inference** |
|---|---|---|
| Goal | Forecast $Y_\text{new}$ | Estimate $\partial Y / \partial X$ |
| Key assumption | Stable data-generating process | Exogeneity / no OVB |
| OLS sufficient? | Often yes | Only under strict conditions |
| Question answered | "What wage would this worker likely earn?" | "What wage *would* this worker earn with one more year of education?" |

Answering causal questions requires additional assumptions and methods
beyond OLS.
:::

The returns-to-education question has motivated decades of empirical
research and methodological innovation: instrumental variables,
regression discontinuity, difference-in-differences. These tools go
beyond this chapter.
<!-- TODO: link to @sec-causalInference once a causal-inference chapter exists. -->

## References {-}

